---
title: "Figure 3 Construction"
author: "Anthony R Cillo"
date: "March 6 2025"
output: md_document
---

## Load packages

```{r}

.libPaths("/ix1/acillo/arc85/packages/Rlibs_sarc028")

library(tidyverse)
library(irlba)
library(ggplot2)
library(patchwork)
library(xlsx)
library(ggplotify)
library(here)

```

## Load data

```{r}

## Read in GeoMx data
geomx_data <- readRDS("../../01_input/SARC028_GeoMx_Expression_and_Metadata.rds")

row_data <- geomx_data$geomx_feature_data
meta_data <- geomx_data$geomx_meta
expr_data <- geomx_data$geomx_expr

```

## Normalization strategy

Rather than directly correcting for AOI surface area, we will attempt to log normalize the protein counts based on the total counts per ROI, which should be correlated with the AOI surface area. 

```{r}

data_expr_fil <- expr_data[!row_data$code_class %in% c("Control","Negative"),]
rownames(data_expr_fil) <- row_data %>%
  filter(!code_class %in% c("Control","Negative")) %>%
  pull(gene)

data_expr_fil <- data_expr_fil %>%
  as_tibble(.,rownames="genes")

log_normalize <- function(x) {
  log(x/sum(x)*1000)
}

data_norm <- apply(data_expr_fil[,2:ncol(data_expr_fil)],2,log_normalize)

```

## PCA with log normalized data 

```{r}

set.seed(123456)

pca_res <- prcomp_irlba(t(data_norm),center=TRUE,scale.=TRUE)
plot(pca_res)

pca_x <- pca_res$x
pca_x <- cbind(meta_data,pca_x)

pca_x %>%
  ggplot(.,aes(x=PC1,PC2,colour=log(aoi_surface_area))) +
    geom_point() +
    theme_bw()

pca_x %>%
  ggplot(.,aes(x=PC1,PC2,colour=segment_name_label)) +
    geom_point() +
    theme_bw()

p2 <- pca_x %>%
  ggplot(.,aes(x=PC1,PC2,colour=timepoint)) +
    geom_point() +
    facet_wrap(~segment_name_label) +
    theme_bw()

```

## DEGs with specific regions

```{r}

data_norm_meta <- rbind(data_norm,t(meta_data)) %>%
  t() %>%
  data.frame() %>%
  as_tibble()

data_norm_meta %>% 
  data.frame() %>%
  select(segment_name_label,timepoint) %>%
  table()

colnames(data_norm_meta)[1:62] <- data_expr_fil$genes

pre_post_degs_segments <- data_norm_meta %>%
  select(-roi_id,-segment,-roi_x_coord,-roi_y_coord,-aoi_surface_area,-aoi_nuclei_count,-slide,-roi,-sample_identifier) %>%
  gather(gene,value,-segment_name_label,-timepoint,-deident_id) %>%
  mutate(value=as.numeric(value)) %>%
  mutate(timepoint=as.factor(timepoint)) %>%
  filter(!segment_name_label=="Tumor") %>% 
  split(.$segment_name_label) %>%
  lapply(.,function(x) {
    x %>%
      split(.$gene) %>%
        map(~wilcox.test(.$value~.$timepoint,data=.)$p.value)
  })
 
degs_sig <- pre_post_degs_segments %>%
  lapply(.,function(x) {
    x[x<0.05]
  })

```

## Heatmaps for distinct subsets - Stromal region

```{r}

stromal_genes <- names(degs_sig$Stroma)
annot_col <- data_norm_meta %>% 
  filter(segment_name_label=="Stroma") %>%
  select(timepoint) %>%
  data.frame()
annot_col$timepoint <- as.factor(annot_col$timepoint)

pid_names <- data_norm_meta %>% 
  filter(segment_name_label=="Stroma") %>%
  select(timepoint,deident_id) %>%
  mutate(pid=paste(deident_id,seq(1,21,1),sep="__")) 

rownames(annot_col) <- pid_names$pid

heat_dat <- data_norm_meta %>% 
  filter(segment_name_label=="Stroma") %>%
  select(-timepoint,-segment_name_label) %>%
  select(data_expr_fil$genes) %>%
  as.matrix() %>%
  apply(.,2,as.numeric) %>% 
  scale() %>%
  t()
colnames(heat_dat) <- pid_names$pid

p3 <- ggplotify::as.ggplot(
  pheatmap::pheatmap(heat_dat,scale="none",annotation_col = annot_col,fontsize = 12,
                   fontsize_row=6,border_color=NA,color=colorRampPalette(colors=c("purple","black","yellow"))(100))
)

```

## Heatmaps for distinct subsets - Full ROI

```{r}

full_roi_genes <- names(degs_sig$`Full ROI`)
annot_col <- data_norm_meta %>% 
  filter(segment_name_label=="Full ROI") %>%
  select(timepoint) %>%
  data.frame()
annot_col$timepoint <- as.factor(annot_col$timepoint)

pid_names <- data_norm_meta %>% 
  filter(segment_name_label=="Full ROI") %>%
  select(timepoint,deident_id) %>%
  mutate(pid=paste(deident_id,seq(1,21,1),sep="__")) 

rownames(annot_col) <- pid_names$pid

heat_dat <- data_norm_meta %>% 
  filter(segment_name_label=="Full ROI") %>%
  select(-timepoint,-segment_name_label) %>%
  select(data_expr_fil$genes) %>%
  as.matrix() %>%
  apply(.,2,as.numeric) %>% 
  scale() %>%
  t()
colnames(heat_dat) <- pid_names$pid

p4 <- ggplotify::as.ggplot(
  pheatmap::pheatmap(heat_dat,scale="none",annotation_col = annot_col,fontsize = 12,
                   fontsize_row=6,border_color=NA,color=colorRampPalette(colors=c("purple","black","yellow"))(100))
)

```

## Plots of DEGs in Full ROI and stroma

```{r}

expr_tumor_stroma <- data_norm_meta %>%
  select(-roi_id,-segment,-roi_x_coord,-roi_y_coord,-aoi_surface_area,-aoi_nuclei_count,-slide,-roi) %>%
  gather(gene,value,-segment_name_label,-timepoint,-deident_id) %>%
  mutate(value=as.numeric(value)) %>%
  mutate(timepoint=as.factor(timepoint)) %>%
  filter(segment_name_label %in% c("Full ROI","Stroma")) %>% 
  split(.$segment_name_label)

degs_sub_list <- degs_sig[3:4]

deg_plots <- mapply(function(x,y) {
  x %>%
    filter(gene %in% names(y)) %>%
    ggplot(.,aes(x=timepoint,y=value,fill=timepoint)) +
    geom_boxplot() +
    theme_bw() +
    facet_wrap(~gene)
},
expr_tumor_stroma,degs_sub_list,SIMPLIFY = F)

```

## Show figure components

```{r}

p2
p3
deg_plots[[2]]
p4
deg_plots[[1]] 

```

## Session info

```{r}

sessionInfo()

```
