---
title: "Integrate NK and T cells from scRNAseq to CosMx"
author: "Anthony R Cillo"
date: "July 17 2024"
output: html_document
---

## Introduction

Here, we will subcluster NK / T cells and B / pDCs to perform refined cell type label transfer.

## Load packages

```{r}

.libPaths("/ihome/acillo/arc85/Rlibs_Mar_2023")

suppressMessages({
    library(Seurat)
    library(tidyverse)
    library(patchwork)
})

```

## Load datasets

```{r}

ser_comb <- readRDS("../03_output/sarc028_cosmx_scrnaseq_integrated_240717.rds")

DimPlot(ser_comb,group.by="initial_cell_annot",label=T)

ser_comb <- ser_comb[,grepl("T_and_NK_cells",ser_comb@meta.data$initial_cell_annot)]

table(ser_comb@meta.data$initial_cell_annot)

DimPlot(ser_comb,group.by="initial_cell_annot")

```

## Create data for joint integration

```{r}

DefaultAssay(ser_comb) <- "RNA"

ser_split <- SplitObject(ser_comb,split.by="data_type")

```

## Combine and integrate data 

```{r}

ser_list <- ser_split

# normalize and identify variable features for each dataset independently
ser_list <- lapply(X = ser_list, FUN = function(x) {
    x <- NormalizeData(x)
    x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 500)
})

# select features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = ser_list)

anchors <- FindIntegrationAnchors(object.list = ser_list, anchor.features = features)

# this command creates an 'integrated' data assay
ser_comb <- IntegrateData(anchorset = anchors)

# specify that we will perform downstream analysis on the corrected data note that the
# original unmodified data still resides in the 'RNA' assay
DefaultAssay(ser_comb) <- "integrated"

# Run the standard workflow for visualization and clustering
ser_comb <- ScaleData(ser_comb, verbose = FALSE)
ser_comb <- RunPCA(ser_comb, npcs = 30, verbose = FALSE)

ElbowPlot(ser_comb)

ser_comb <- RunUMAP(ser_comb, reduction = "pca", dims = 1:12)
ser_comb <- FindNeighbors(ser_comb, reduction = "pca", dims = 1:12)
ser_comb <- FindClusters(ser_comb, resolution = c(0.3,0.5,0.7,1))

```

## Check results 

```{r}

DimPlot(ser_comb,group.by="data_type")

p1 <- FeaturePlot(ser_comb,c("FOXP3","CD8A","FCGR3A","CD4"))

p2 <- DimPlot(ser_comb,group.by="integrated_snn_res.0.5",label=T)

p3 <- DimPlot(ser_comb,group.by="integrated_snn_res.1",label=T)

p1 + p2

p3

Idents(ser_comb) <- "integrated_snn_res.1"

cluster7 <- FindMarkers(ser_comb,ident.1="7")
cluster8 <- FindMarkers(ser_comb,ident.1="8")

cluster7 %>%
    filter(avg_log2FC>0) %>%
    filter(p_val_adj<0.05)

cluster8 %>%
    filter(avg_log2FC>0) %>%
    filter(p_val_adj<0.05)

meta <- ser_comb@meta.data %>%
    mutate(refined_lymph_cell_annot=recode(integrated_snn_res.1,
                                   `0` = "CD8pos_T_cell",
                                   `1` = "CD4pos_Tconv",
                                   `2` = "CD4pos_Tconv",
                                   `3` = "CD4pos_Tconv",
                                   `4` = "CD4pos_Tconv",
                                   `5` = "CD8pos_T_cell",
                                   `6` = "CD8pos_T_cell",
                                   `7` = "NK_cell",
                                   `8` = "CD8pos_T_cell",
                                   `9` = "CD8pos_T_cell",
                                   `10` = "CD4pos_Tconv",
                                   `11` = "CD4pos_Tconv",
                                   `12` = "CD4pos_Treg",
                                   `13` = "CD8pos_T_cell"
                                                                     )
           )

meta$refined_lymph_cell_annot <- as.character(meta$refined_lymph_cell_annot)

table(meta$refined_lymph_cell_annot,useNA="always")

ser_comb@meta.data$refined_nk_t_cell_annot <- meta$refined_lymph_cell_annot

DimPlot(ser_comb,group.by="refined_nk_t_cell_annot")

refined_nk_t_results <- ser_comb

```

## Refine plasmablast and pDC cells

```{r}

ser_comb <- readRDS("../03_output/sarc028_cosmx_scrnaseq_integrated_240717.rds")

ser_comb <- ser_comb[,grepl("Plasmablasts/pDCs",ser_comb@meta.data$initial_cell_annot)]

table(ser_comb@meta.data$initial_cell_annot)

DimPlot(ser_comb,group.by="initial_cell_annot")

DimPlot(ser_comb,group.by="data_type")

FeaturePlot(ser_comb,c("MZB1","IL3RA","JCHAIN"))

```

## Combine and integrate data 

```{r}

DefaultAssay(ser_comb) <- "RNA"

ser_comb <- ser_comb %>%
    NormalizeData(.) %>%
    FindVariableFeatures(.) %>%
    ScaleData(.) %>%
    RunPCA(.)

ElbowPlot(ser_comb)

ser_comb <- ScaleData(ser_comb, verbose = FALSE)
ser_comb <- RunPCA(ser_comb, npcs = 30, verbose = FALSE)

ElbowPlot(ser_comb)

ser_comb <- RunUMAP(ser_comb, reduction = "pca", dims = 1:15)
ser_comb <- FindNeighbors(ser_comb, reduction = "pca", dims = 1:15)
ser_comb <- FindClusters(ser_comb, resolution = c(0.3,0.5,0.7,1))

```

## Refine Plasmablasts and pDCs

```{r}

DimPlot(ser_comb,group.by="RNA_snn_res.0.3",label=T)

Idents(ser_comb) <- "RNA_snn_res.0.3"

degs <- FindAllMarkers(ser_comb)

degs %>%
    filter(p_val_adj<0.05) %>%
    filter(avg_log2FC>0) %>%
    split(.$cluster) %>%
    lapply(.,function(x) x %>% arrange(desc(avg_log2FC)) %>% head(n=10))

FeaturePlot(ser_comb,c("MS4A1","IL3RA"))

b_pdc_meta <- ser_comb@meta.data %>%
    mutate(refined_b_pdc_annot=ifelse(RNA_snn_res.0.3 %in% c(4,1,7,3),"Plasmablasts","pDCs")) %>%
    mutate(refined_b_pdc_annot=ifelse(RNA_snn_res.0.3 %in% c(0,5),NA,refined_b_pdc_annot))

table(b_pdc_meta$refined_b_pdc_annot,useNA="always")

```


## Recombine T cell subsets original data

```{r}

## Reload original data
ser_comb <- readRDS("../03_output/sarc028_cosmx_scrnaseq_integrated_240717.rds")

meta_orig <- ser_comb@meta.data %>%
    as_tibble(.,rownames="cb")

meta_new1 <- refined_nk_t_results@meta.data %>%
    as_tibble(.,rownames="cb") %>%
    select(cb,refined_nk_t_cell_annot)

meta_new2 <- b_pdc_meta %>%
    as_tibble(.,rownames="cb") %>%
    select(cb,refined_b_pdc_annot)

meta_all <- left_join(meta_orig,meta_new1,by="cb") %>%
    left_join(.,meta_new2,by="cb") %>%
    mutate(final_cell_annot=ifelse(is.na(refined_nk_t_cell_annot),
                                           as.character(initial_cell_annot),
                                           refined_nk_t_cell_annot)
           ) %>%
    mutate(final_cell_annot=ifelse(is.na(refined_b_pdc_annot),
                                           as.character(final_cell_annot),
                                           refined_b_pdc_annot)
           )


table(meta_all$final_cell_annot)

ser_comb@meta.data$final_cell_annot <- meta_all$final_cell_annot

ser_comb <- ser_comb[,!ser_comb@meta.data$final_cell_annot=="Plasmablasts/pDCs"]

DimPlot(ser_comb,group.by="final_cell_annot")

```

## Save results 

```{r}

saveRDS(ser_comb,file="../03_output/sarc028_cosmx_scrnaseq_final_cell_annot_ser_240717.rds")

```

## Plotting 

```{r}

ser_split <- SplitObject(ser_comb,split.by="data_type")

p1 <- DimPlot(ser_split[[1]],group.by="final_cell_annot") +
    ggtitle("scRNAseq") +
    theme(legend.position="none")
p2 <- DimPlot(ser_split[[2]],group.by="final_cell_annot") +
    ggtitle("Nanostring CosMx") +
    theme(legend.position="none")

p1 / p2 

```

